%!TEX encoding = UTF-8 Unicode

%% Thesis cover pages
%%
%% Based on the example.tex file provided by Tomás Oliveira e Silva
%%
%% Modified by João Paulo Barraca <jpbarraca@ua.pt>
%% Modified by Filipe Manco <filipe.manco@ua.pt>

% Remove 'coverdepname' parameter to remove department name from cover pages
% Add 'final' parameter to remove "DOCUMENTO PROVISÓRIO" from cover page
\documentclass[11pt,twoside,a4paper,final]{report}

%%% This file is a template for the cover of your
%%%  thesis.
%%% Lines like this one, starting with %%% , are
%%%  instructions. Search them to know where and
%%%  how to edit this file. After editing the
%%%  document you can delete them.
%%% Items marked with OPTIONAL can be removed.
%%% Most of the lines you will have to change
%%%  already have examples to help you understand.
%%% The complete list of options and commands can
%%%  be found in uaThesisTemplate.sty


%%% Select your area of study and department from
%%%  the following lists and replace them in the
%%%  \usepackage declaration ([area,department]).
%%%
%%% | area        |   | department |
%%% ---------------   --------------
%%% | arts        |   | dao        |
%%% | sciences    |   | dbio       |
%%% | education   |   | dcspt      |
%%% | economy     |   | deca       |
%%% | engineering |   | degeit     |
%%% | humanities  |   | de         |
%%% | health      |   | deti       |
%%% |             |   | demac      |
%%% |             |   | decivil    |
%%% |             |   | dem        |
%%% |             |   | fis        |
%%% |             |   | dlc        |
%%% |             |   | dmat       |
%%% |             |   | dq         |
%%% |             |   | dcm        |
%%% |             |   | geo        |
%%% ---------------   --------------
%% Import the UA thesis template
\usepackage[engineering,deti]{uaThesisTemplate}


%% Optional packages for utf8 encoding and T1 fonts
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

%% For prettier pages
\usepackage{lmodern}

%% We will use english and portuguese
\usepackage[english]{babel}

\usepackage{xspace}% used by \sigla

%%% This package is only used to create a dummy
%%%  text for the abstract. You can remove the
%%%  following line in the final document.
\usepackage{blindtext}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  START
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Edit thesis year
%% Thesis Year
\ThesisYear{2021}


%%% Uncomment one of the following
%\DocumentTypeThesis
\DocumentTypeDissertation

%%% Edit author's name
%% Author's full name
\Author{André  Silva\\Pedrosa}

%%% Edit thesis title in Portuguese
%% Thesis Mandatory Title in Portuguese
\Title{\selectlanguage{portuguese}Desenvolvimento de uma arquitetura escalável para extrair metainformação de bases de dados médicas distribuídas}

%%% OPTIONAL: you can write your thesis title in
%%%  other languages. You can repeat the following
%%%  line as many times as you want, or remove it
%%%  if you only want the portuguese title.
%% Optional Titles in additional languages.
\Title{\selectlanguage{english}Development of a scalable architecture to extract metadata from distributed medical databases}

%%% OPTIONAL: add a quotation if you like.
%% Optional Quotation
\Quote{\selectlanguage{english}You can't have happiness without pain, you need a little bit of rain to have a little bit of rainbow}{Felix Lengyel}

%%% Edit your supervisor's information.
%%%
%%% \Supervisor{name}{category}
%%%
%%% name   : Supervisor's name
%%% categoy: One of the following list
%%%
%%% | category               |
%%% --------------------------
%%% | catedrático            |
%%% | catedrático convidado  |
%%% | associado c/ agregação |
%%% | associado              |
%%% | auxiliar               |
%%% | auxiliar convidado     |
%%% --------------------------
%% Supervisor information
\Supervisor{José Luís Guimarães Oliveira}{catedrático}

%%% OPTIONAL: you can add a second supervisor.
%%%
%%% \SecondSupervisor{name}{category}
%%%  {[preposition: (at, in, no, da)] department}
%%%  {[preposition: (at, in, do, na)] school}
%%%
%%% name      : See above.
%%% category  : See above.
%%% department: Can be a department, faculty or
%%%             other; Choose the preposition
%%%             accordingly.
%%% school    : University/Scholl name; Choose the
%%%             preposition accordingly.
%% Optional SecondSuperviser information
%\SecondSupervisor{João Rafael Duarte de Almeida}{Investigador}{do Instituto de Engenharia Electrónica e Telemática de Aveiro}{}

%%% Edit your degree information.
%%%
%%% \Degree{title}{name}
%%%
%%% title: Degree title (Mestre or Doutor)
%%% name : Degree name
%% Degree title and name
\Degree{Mestre}{Engenharia de Informática}

%%% OPTIONAL: edit grant support texts where
%%%  appropriate. Maximum two.
%%%
%%% \GrantText{grant1}{grant2}
%% Optional Grant support texts
%\GrantText{Texto Apoio financeiro do POCTI no âmbito do III Quadro Comunitário de Apoio.}
%          {Texto Apoio financeiro da FCT e do FSE no âmbito do III Quadro Comunitário de Apoio.}

%%% OPTIONAL: Add a dedication text.
%% Optional Dedication text
%\Dedication{Dedico este trabalho à minha esposa e filho pelo incansável apoio.}

%%% OPTIONAL: Add an acknowledgement text
%% Optional Acknowledgement text
\Acknowledgements{%
Começo por agradecer a todas as pessoas que conheci por meio da minha estadia na residência.
Dos veteranos aos caloiros um grande obrigado por me darem uma razão de continuar a trabalhar e por criarem um ambiente de família.
\\\\
Aos meu orientador, o professor José Luís Oliveira, e coorientador, João Almeida, por acreditarem no meu trabalho e não desistirem de mim, para além das várias vezes que possa não ter atingido as suas expectativas.
\\\\
Agradeço aos vários amigos e amigos que fui conhecendo ao longo do meu percurso académico da Universidade de Aveiro.
Em especial, obrigado aos fundadores do grupo SASA LELE, Filipe Pires e João Alegria, por suportarem o meu feitio e por estarem ao meu lado durante o nosso curso.
\\\\
À malta de Leiria que para além de me ter descolado para um local distante, continuarem a receber-me como um amigo chegado.
Agradeço especialmente ao Leandro pela paciência e preocupação.
\\\\
Por último, mas não menos importante, queria agradecer à minha família por estarem lá nos meus altos e baixos e por me darem força para eu seguir em frente.
}

%%% Edit the keywords and abstract in Portuguese
%%% Don't forget to remove the \blindtext[3]
%%%  declaration.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation
%% Mandatory Keywords and Abstract in Portuguese
\Abstract{Palavras Chave}{meta-dados, registo eletrónico de pacientes, EHDEN, OHDSI, atualização, automatização, extração}
         {Resumo}{\selectlanguage{portuguese}%
Para realizar estudos médicos, tais como o impacto de um fármaco numa determinada população, os investigadores precisam de acesso a dados reais de organizações médicas, tais como hospitais, para que os seus estudos tenham resultados fiáveis. %
No entanto, para os investigadores, este é um processo difícil e demorado até conseguirem encontrar o conjunto de dados que melhor se adapta ao seu estudo, dado que muitas vezes não têm acesso direto aos dados reais. %
Para ajudar neste processo, existem plataformas centralizadas que descrevem o conteúdo dos conjuntos de dados através de meta-dados. %
Atualmente, em muitas destas plataformas, estes meta-dados são atualizados manualmente, um processo lento e tedioso, o que leva a que se tornem muito facilmente desatualizados. %
\\%
O objetivo deste trabalho é desenvolver um sistema que extrai automaticamente meta-dados diretamente dos dados reais das bases de dados e os envia para várias plataformas que estão a expor meta-dados de modo a que estes se mantenham atualizados. %
O sistema pretende trabalhar em bases de dados ligadas ao grupo EHDEN (European Health Data and Evidence Network), que pretende criar uma rede federada de bases de dados médicas na Europa, seguindo os mesmos princípios e ferramentas utilizados pela comunidade OHDSI (Observational Health Data Sciences and Informatics). %
A extração de meta-dados é feita por agentes que são instalados juntamente com o sistema de produção das bases de dados. %
Os meta-dados são então enviados para um sistema de gestão de meta-dados, construído sobre uma framework de fluxo de dados chamada Kafka. %
Este sistema de gestão é composto por um conjunto de componentes que foram desenvolvidos de acordo com uma filosofia de microsserviços, tendo sempre em mente a possibilidade de escalar cada um deles.%
}

%%% OPTIONAL: you can write the keywords and
%%%  abstract in other languages. You can repeat
%%%  the following lines as many times as you want
%%%  or remove them if you only want portuguese.
%%% Don't remove the \selectlanguage statement or you will end with incorrect hyphenation.
%%% You can modify the language to any other supported by Latex
%% Optional Keywords and Abstract in additional languages.
\Abstract{Keywords}{metadata, electronic health records, EHDEN, OHDSI, update, automation, extraction}
         {Abstract}{\selectlanguage{english}%
To conduct medical studies, such as the impact of a drug on a certain population, researchers need access to real data from medical organizations such as hospitals for their studies to have reliable results. %
However, for researchers, this is a difficult and time-consuming process until they can find the dataset that best fits their study, as they often do not have direct access to the actual data. %
To help in this process, there are centralized platforms that describe the content of datasets through metadata. %
Currently, in many of these platforms, this metadata is updated manually, a slow and tedious process, which leads to it becoming outdated very easily. %
\\%
The goal of this work is to develop a system that automatically extracts metadata directly from databases' actual data and sends it to various platforms that are exposing metadata so that they are kept up to date. %
The system intends to work on databases linked to the EHDEN (European Health Data and Evidence Network) group, which intends to create a federated network of medical databases in Europe, following the same principles and tools used by the OHDSI (Observational Health Data Sciences and Informatics) community. %
Metadata extraction is done by agents that are installed along with the production system of the databases. %
The metadata is then sent to a metadata management system, built on top of a data streaming framework called Kafka. %
This management system is composed of a set of components that were developed according to a philosophy of microservices, always keeping in mind the possibility of scaling each one.%
}

%%% Edit the Jury President information.
%% First Member will be the President of the Jury
\JuriMember{Prof. Doutor Joaquim Arnaldo Carvalho Martins}{Professor Catedrático, Universidade de Aveiro}

%%% Repeat the above declaration for additional
%%%  members of the jury as many times as necessary.
%% Additional members of the Jury
\JuriMember{Prof. Doutor Paula Alexandra Gomes da Silva}{Professora Auxiliar, Universidade de Coimbra - Faculdade de Ciências e Tecnologia}
\JuriMember{Prof. Doutor José Luis Guimarães Oliveira}{Professor Catedrático, Universidade de Aveiro}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%       DEFINE THESIS COVER FIELDS
%%                  END
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\begin{document}

%%% By default we consider portuguese to be the
%%%  main language (the last one in the usepackage
%%%  declaration). If you're writing your thesis
%%%  in english uncomment the following line.
%%% Set the proper language or you will end with incorrect hyphenation!
\selectlanguage{english}

\MakeCover

\end{document}
